---
title: "Simulating random effects"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("lme4")
library("dplyr")
library("ggplot2")
```

Simulate a dataset with 5 individuals, each one containing three replicates.

Specify desired between-individual variance and measurement noise variance
```{r}
var_individual = 0.5
var_noise = 0.25
```

```{r}
#Specify the number of replicates per individual
replicate_counts = c(10,10,10,10,10)

#Sample individual means from the normal distribution with variance var_individual
individual_means = rnorm(5, mean = 0, sd = sqrt(var_individual))
ind_means = dplyr::data_frame(individual_mean = individual_means, 
                              individual = paste0("ind_", c(1:5)))

#Add replicates and sample measurement noise
data_balanced = dplyr::data_frame(individual = paste0("ind_", rep(c(1,2,3,4,5), times = replicate_counts))) %>%
  dplyr::left_join(ind_means, by = "individual") %>%
  #Sample measurement noise from the normal distribution with variance var_noise
  dplyr::mutate(noise = rnorm(length(individual), mean = 0, sd = sqrt(var_noise))) %>%
  dplyr::mutate(expression = individual_mean + noise) %>%
  dplyr::mutate(sample_index = c(1:length(individual)))
data_balanced
```

Theoretical (population) variance explained by individual
```{r}
var_individual/(var_individual + var_noise)
```

Simulated sample variance explained by individual
```{r}
var(individual_means)/(var(individual_means) + var(data_balanced$noise))
```

Make a plot of the raw data
```{r}
ggplot(data_balanced, aes(x = sample_index, y = expression, color = individual)) + 
  geom_point() + 
  facet_grid(~individual, scales = "free_x")
```

Specify a helper function that extract % variance explained from the linear mixed model
```{r}
#' Calculate the proportion of variance explaned by different factors in a lme4 model
varianceExplained <- function(lmer_model){
  variance = as.data.frame(lme4::VarCorr(lmer_model))
  var_percent = dplyr::mutate(variance, percent_variance = vcov/sum(vcov)) %>% 
    dplyr::select(grp, percent_variance) %>% 
    dplyr::mutate(type = "gene")
  var_row = tidyr::spread(var_percent, grp, percent_variance)
  return(var_row)  
}
```

Fit a linear mixed model to the dataset estimate variance explained
```{r}
model = lmer(expression ~ (1|individual), data_balanced)
varianceExplained(model)
```

To the same with a standard fixed effect linear model
```{r}
model_fixed = lm(expression ~ individual, data_balanced)
#Estimate variance explained
summary(model_fixed)$adj.r.squared
```

Extract model coefficients for random effects
```{r}
coefs = coef(model)$individual
coefs_df = dplyr::data_frame(individual = rownames(coefs), coef = coefs[,1], type = "lme4")
```

Estmate individual means
```{r}
ind_means = dplyr::group_by(data_balanced, individual) %>% 
  dplyr::summarize(coef = mean(expression), type = "lm")
```

```{r}
ggplot(data_balanced, aes(x = sample_index, y = expression, color = individual)) + 
  geom_point() + 
  facet_grid(~individual, scales = "free_x") + 
  geom_hline(data = coefs_df, aes(yintercept = coef, linetype = type)) +
  geom_hline(data = ind_means, aes(yintercept = coef, linetype = type))
```



# Application on an unbalanced dataset

```{r}
#Specify the number of replicates per individual
replicate_counts = c(20,15,13,1,1)

#Sample individual means from the normal distribution with variance var_individual
individual_means = rnorm(5, mean = 0, sd = sqrt(var_individual))
ind_means = dplyr::data_frame(individual_mean = individual_means, 
                              individual = paste0("ind_", c(1:5)))

#Add replicates and sample measurement noise
data_unbalanced = dplyr::data_frame(individual = paste0("ind_", rep(c(1,2,3,4,5), times = replicate_counts))) %>%
  dplyr::left_join(ind_means, by = "individual") %>%
  #Sample measurement noise from the normal distribution with variance var_noise
  dplyr::mutate(noise = rnorm(length(individual), mean = 0, sd = sqrt(var_noise))) %>%
  dplyr::mutate(expression = individual_mean + noise) %>%
  dplyr::mutate(sample_index = c(1:length(individual)))
data_unbalanced
```

Visualise the data
```{r}
ggplot(data_unbalanced, aes(x = sample_index, y = expression, color = individual)) + 
  geom_point() + 
  facet_grid(~individual, scales = "free_x")
```

Simulated variance explained
```{r}
var(individual_means)/(var(individual_means) + var(data_unbalanced$noise))
```


```{r}
model = lmer(expression ~ (1|individual), data_unbalanced)
varianceExplained(model)
```

```{r}
model_fixed = lm(expression ~ individual, data_unbalanced)
#Estimate variance explained
summary(model_fixed)$adj.r.squared
```

Extract model coefficients for random effects
```{r}
coefs = coef(model)$individual
coefs_df = dplyr::data_frame(individual = rownames(coefs), coef = coefs[,1], type = "lme4")
model_df = dplyr::left_join(data_unbalanced, coefs_df, by = "individual")
```

Estmate individual means
```{r}
ind_means = dplyr::group_by(data_unbalanced, individual) %>% 
  dplyr::summarize(coef = mean(expression), type = "lm")
```

```{r}
ggplot(data_unbalanced, aes(x = sample_index, y = expression, color = individual)) + 
  geom_point() + 
  facet_grid(~individual, scales = "free_x") + 
  geom_hline(data = coefs_df, aes(yintercept = coef, linetype = type)) +
  geom_hline(data = ind_means, aes(yintercept = coef, linetype = type))
```

