---
title: "Exploring gene expression data"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
suppressWarnings(suppressMessages(library("SummarizedExperiment")))
suppressWarnings(suppressMessages(library("dplyr")))
suppressWarnings(suppressMessages(library("DESeq2")))
suppressWarnings(suppressMessages(library("gplots")))
suppressWarnings(suppressMessages(library("ggplot2")))
```

First, we need to load the libraries:

```{r}
library("SummarizedExperiment")
library("DESeq2")
library("dplyr")
library("gplots")
library("ggplot2")
```

## Importing the data

First, download the full dataset from the course [website](https://zenodo.org/records/10805436/files/RNA_SummarizedExperiment.rds?download=1).

Now you can import the dataset into R using the following command:

```{r}
	dataset = readRDS("../data/RNA_SummarizedExperiment.rds")
```


Here, `dataset` is an [SummarizedExperiment](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html)  object. The key advantage of using SummarizedExperiment class is that the gene expression read count matrix is always stored together with the accompanying gene and sample metadata. Consequently, when you subset the data, SummarizedExperiment ensures that the metadata is always synchronised with the expression data itself. I strongly recommend you read the  [SummarizedExperiment](https://bioconductor.org/packages/release/bioc/vignettes/SummarizedExperiment/inst/doc/SummarizedExperiment.html) tutorial to learn how to perform basic operations with these objects.

You can see the summary of the dataset by just printing it out:
```{r}
	dataset
```
The `## dim` line tells to that this dataset has 35033 genes and 336 samples. 

You can view the sample metadata (first 5 rows and columns) using the `colData()` function:
```{r}
	colData(dataset)[1:5,1:5]
```
And you can do the same for gene metadata:
```{r}
	rowData(dataset)[1:5,]
```

Finally, to view the raw read counts, you can use the assays command:
```{r}
	assays(dataset)
```

```{r}
	assays(dataset)$counts[1:5,1:5]
```

# Subsetting
To illustrate the power of SummarizedExperiment, let's take a random subset of the data containing eight random donors. First, let's identify the donors:
```{r}
	set.seed(1)
  donors_subset = colData(dataset)$donor %>% unique() %>% sample(8)
  donors_subset
```
Note that I used the `%>%` (pipe) operator from the [dplyr](https://dplyr.tidyverse.org/) pacakge. The last command is exactly equivalent (but more readable) as the following:
```{r}
	set.seed(1)
  donors_subset = sample(unique(colData(dataset)$donor),8)
  donors_subset
```

Now we can just subset the original dataset:
```{r}
data_subset = dataset[,dataset$donor %in% donors_subset]
data_subset
```
Note the that we now have 8 donors x 4 conditions = 32 samples.

# Expolratory data analysis
We will use the [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html) package to analyse the RNA-seq data. DESeq2 also has [excellent documentation](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html). 

First, let's convert the SummarizedExperiment object into a DESeq object:
```{r}
dds <- DESeqDataSet(data_subset, design = ~ condition_name)
dds
```

Next, let's perform variance stabilising normalisation:
```{r}
vsd <- vst(dds, blind=FALSE)
```
This is what the data looks like after normalisation:
```{r}
  assay(vsd)[1:5, 1:5]
```
Now, let's etract 10,000 most highly expressed genes
```{r}
  highly_expressed = names(sort(rowMeans(assay(vsd)), decreasing = TRUE)[1:10000])
  vsd = vsd[highly_expressed,]
```

The simplest exploratory analysis is to look at the heatmap pairwise sample-to-sample correlations:
```{r}
cor_matrix = cor(assay(vsd), method = "spearman")
gplots::heatmap.2(cor_matrix)
```

Second approach is to perform PCA. DESeq2 has a built-in function for that. Note that by default the plotPCA function uses only the top 500 genes with highest variance between samples.
```{r}
  plotPCA(vsd, intgroup=c("condition_name"))
```

You can modify this behaviour be changing the ntop parameter:
```{r}
  plotPCA(vsd, intgroup=c("condition_name"), ntop = 10000)
```

Alternatively, you can do it manually using prcomp and ggplot2 functions. The advantage of the manual approach is that you can customise the plot more, including adding names for individual samples. You can also decide to include all of the genes in the analysis. 
```{r}
  pca = prcomp(t(assay(vsd)), center = TRUE, scale. = TRUE)
  pca_matrix = dplyr::mutate(as.data.frame(pca$x), sample_id = rownames(pca$x))
  ggplot(pca_matrix, aes(x = PC1, y = PC2, label = sample_id)) + geom_text() + geom_point()
```
## Differential expression

For simplicity, let's first look at only two conditions - naive and interferon-gamma stimulation:

```{r}
naive_IFNg = data_subset[,data_subset$condition_name %in% c("naive", "IFNg")]
```

Make DEseq2 object
```{r}
dds <- DESeqDataSet(naive_IFNg, design = ~ condition_name)
dds$condition_name = relevel(dds$condition_name, ref = "naive")
```

Remove genes that have very low absolute expression levels
```{r}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
```

#Run DEseq2
```{r}
dds <- DESeq(dds)
ifng_results <- results(dds)
```

Make a histogram of the p-values
```{r}
hist(ifng_results$pvalue)
```

#Identify genes that are signficantly differentially expressed:
```{r}
filtered_genes = dplyr::mutate(as.data.frame(ifng_results), gene_id = rownames(ifng_results)) %>% 
  dplyr::as_tibble() %>% 
  dplyr::filter(!is.na(padj)) %>% #Remove NA p-values
  dplyr::filter(padj < 0.01) %>% #Filter according to FDR
  dplyr::filter(abs(log2FoldChange) > 1) %>% #Filter according to fold-change
  dplyr::arrange(-log2FoldChange) #Sort by fold change
filtered_genes
```

Save the results to disk:
```{r}
write.table(filtered_genes, "naive_vs_IFNg_gene_list.txt", sep = "\t", quote = FALSE, row.names = FALSE)
```


#Now, let's repeat the same analysis for naive vs Salmonella condition

```{r}
naive_SL1344 = data_subset[,data_subset$condition_name %in% c("naive", "SL1344")]
dds <- DESeqDataSet(naive_SL1344, design = ~ condition_name)
dds$condition_name = relevel(dds$condition_name, ref = "naive")

#Keep expressed genes
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

#Run DEseq2
dds <- DESeq(dds)
salmonella_results <- results(dds)
```

```{r}
#Now, let compare the log2FoldChanges
lfc_SL1344 = dplyr::mutate(as.data.frame(salmonella_results), gene_id = row.names(salmonella_results)) %>% 
  dplyr::as_tibble()
lfc_IFNg = dplyr::mutate(as.data.frame(ifng_results), gene_id = row.names(ifng_results)) %>% 
  dplyr::as_tibble()
joint_lfc = dplyr::inner_join(lfc_IFNg, lfc_SL1344, by = "gene_id")

ggplot(joint_lfc, aes(x = log2FoldChange.x, y = log2FoldChange.y)) + geom_point() +
  xlab("naive vs IFNg") + 
  ylab("naive vs Salmonella")

cor.test(joint_lfc$log2FoldChange.x, joint_lfc$log2FoldChange.y)
```

